% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

\chapter*{Acknowledgments}
\markboth{Acknowledgments}{}

First of all, I'd like to thank my advisor Joel P. Arrais, for all the support he gave me throughout this project, despite all the incredible things that were happening in his life, such as the birth of son, to whom I give my best wishes. He really did handle me a great deal of responsibility and liberty to investigate what I truly wanted, in the subject that I wanted, and for that I am extremely thankful.

I want to thank Conceição Ega and Hugo Froufe from UC-BIOTECH that kindly shared the dataset used in this project with me, and helped answering so many of the questions I had as of the start of this project.

Junior Enterprise for Science and Technology was one of the most surprisingly positive experiences I had the pleasure to be a part of during my time in this University. I want to thank for the incessant discussion of the most random subjects one can think of, for all the shared knowledge, for letting me feel truly welcome as I am, and above all, for being such a bunch of amazing people that keeps growing every year.

To all my friends for the company through these years, for all the laughs, shared moments, work and all nighters. Thank you for being there, sharing joy and pain, for being you. I do really feel like I have a great number of true friends, who have proven so countless times, and I hope to return to you, as much as you give to me.

Wee Ana, there is simply no way of putting on paper how much your presence as positively impacted my life. Thank you for putting up with me, you really do rock my world.

Finally to my father, my mother, my brothers, my uncle and my grandmother, for giving your all in these last tough times, and still having the strength left to be so present as you are in my life. I am really proud to be able to call you family.